
---
title: MALTEM school absenteeism overview
author: 
  - Laia Cirera
  - Joe Brew
  - Elisa Sicuri 
fig_height: 4
fig_width: 4
header-includes: \usepackage{longtable}
highlight: zenburn
latex_engine: xelatex
output:
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
tables: yes
toc: yes
---

```{r, echo = FALSE, warning = FALSE, message = FALSE, comment = NA, error= FALSE }
# No scientific notation
options(scipen=999)

#####
# PACKAGES
#####
library(dplyr)
library(xtable)
library(ggplot2)
library(knitr)
library(png)
library(grid)
library(extrafont)
library(tidyr)
library(gridExtra)
library(maptools)
library(rgeos)
library(ggmap)
library(ggrepel)

loadfonts(device = 'postscript', quiet = TRUE)

# BASIC KNITR OPTIONS
opts_chunk$set(comment = NA, echo = FALSE, warning = FALSE, message = FALSE, error = FALSE, fig.height = 4, fig.align='center')
options(xtable.comment = FALSE)
```

```{r, results='hide'}
# GET THEME FOR GGPLOTS
source('theme.R')
# READ IN DATA
source('read_and_clean_all.R')
```

```{r, fig.align='center', fig.height = 3}
# TITLE MAP
ggplot() +
  coord_map() +
  geom_polygon(data = moz3_fortified,
       aes(x = long, y =lat, group = group), fill = 'grey', color = 'white') +
    geom_polygon(data = maputo_fortified,
       aes(x = long, y =lat, group = group), fill = 'darkred', color = NA, alpha = 0.6) +
  theme_tufte() +
  theme(axis.ticks.length = unit(0.001, "mm")) + labs(x=NULL, y=NULL) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      # legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank()) 
```

Full code at https://github.com/joebrew/maltem_school_absenteeism.

\newpage


# Introduction  

This document presents a largely visual overview of intitial absenteeism data related to the MALTEM program in Magude and Manhiça, Mozambique. Note that we do not present "findings" or "results", but rather general overviews. 

This report is programmed as a "dynamic document"; it can reproduced automatically with updatd data at any point in time.


## Schools' locations

What follows is a basic map of each schools' geographic location.

```{r}
cols <- c('darkorange', 'darkgreen')
ggplot() +
  geom_polygon(data = combined,
       aes(x = long, 
           y = lat,
           group = group,
           fill = District),
       color = 'white',
       alpha = 0.6,
       size = 0.1) +
  coord_map() +
  theme_bw() +
  coord_map() +
  scale_fill_manual(name = '',
                    values = cols) +
  xlab('') +
  ylab('') +
  theme_maltem() +
  theme(axis.ticks.y = element_blank(),axis.text.y = element_blank(), # get rid of x ticks/text
        axis.ticks.x = element_blank(),axis.text.x = element_blank(), # get rid of y ticks/text
        plot.title = element_text(lineheight=.8, face="bold", vjust=1),
        panel.border = element_blank())  +
  geom_point(data = schools,
             aes(x = lng,
                 y = lat),
             alpha = 0.6,
             color = 'darkred') +
  ggrepel::geom_text_repel(data = schools,
             aes(x = lng,
                 y = lat,
                 label = gsub(' ', '\n', SCHOOL_NAME)),
             size = 1.5, 
             alpha = 0.5,
             color = 'darkred') 
```

## Absenteeism by school

The rate of absenteeism for all schools apparently fell from 2015 to 2016. This is not due to a true decline in absenteeism, but rather due to a change in collection techniques. Whereas 2015 had reliable "denominator" data (ie, which days class was held, which days were holidays), 2016 had unreliable "denominator" data.

Though this is problematic, we take comfort in the fact that there is no reason to assume that the "incorrectness" of the "denominator" was any worse for the intervention (Magude) area than in the control (Manhiça) area. In other words, since the true absenteeism rate cannot be ascertained, we can instead rely on a comparison of the (randomly / equally under-estimated) rates.

The below chart shows the extent of the problem.

```{r}
temp <- 
  attendance %>%
  mutate(year = as.numeric(format(date, '%Y'))) %>%
  mutate(school = substr(combined_number, 4,5)) %>%
  group_by(year, school) %>%
  summarise(absences = length(which(absent)),
            eligibles = n()) %>%
  mutate(absenteeism_rate = absences / eligibles * 100)
temp$year <- factor(paste0('Year: ', temp$year))

cols <- c('darkgreen', 'darkorange')
ggplot(data = temp,
       aes(x = school,
           y = absenteeism_rate,
           group = year,
           fill = year)) +
  geom_bar(stat = 'identity', 
           position = 'dodge',
           alpha = 0.7) +
  scale_fill_manual(name = '',
                    values = cols) +
  theme_maltem() +
  xlab('School') +
  ylab('Absenteeism rate') +
  ggtitle('Differential absenteeism rate by year',
          'Due to undercollection in 2016')
```

## Absenteeism over time by district

We want to compare Magude (intervention) with Manhiça (control). Due to the aforementioned problems with the "denominators", the comparison is difficult to visualize.

```{r}
cols <- c('darkorange', 'darkgreen')

temp <- 
  attendance %>%
  mutate(district = as.numeric(substr(combined_number, 1, 2)))%>%
  mutate(district = ifelse(district == 1, 'Manhiça', 
                           ifelse(district == 2, 'Magude', NA))) %>%
  group_by(date, district) %>%
    summarise(absences = length(which(absent)),
            eligibles = n()) %>%
  mutate(absenteeism_rate = absences / eligibles * 100) %>%
  rename(`Students under observation` = eligibles)

ggplot(data = temp,
       aes(x = date,
           y = absenteeism_rate,
           color = district)) +
  geom_point(alpha = 0.3,
             aes(size = `Students under observation`)) +
  geom_line(alpha = 0.3) +
  geom_smooth(aes(weight = `Students under observation`)) +
  scale_color_manual(name = 'District',
                     values = cols) +
  theme_maltem() +
  xlab('Date') +
  ylab('Absenteeism rate') +
  ggtitle('Smoothed absenteeism rate by district') +
  ylim(0,25) +
  geom_vline(xintercept = as.numeric(as.Date(c('2015-11-15', '2016-01-15'))), lty = 2, alpha = 0.6, color = 'darkred') +
  theme(legend.position = 'bottom')
```

To make a more accurate comparison, we (a) assume that the errors in the "denominators" are randomly distributed by district, and then (b) chart a relative rate of absence (rather than the "true" rate). The below chart 

```{r}
x <- temp %>% filter(district == 'Manhiça') %>%
    rename(man = absenteeism_rate,
           man_students = `Students under observation`) %>%
    dplyr::select(date, man, man_students)
y <- temp %>% filter(district == 'Magude') %>%
    rename(mag = absenteeism_rate,
           mag_students = `Students under observation`) %>%
    dplyr::select(date, mag, mag_students)
temp2 <- full_join(x, y) %>%
  mutate(mag_to_man = (mag / man)  * 100) %>%
  mutate(`Students under observation` = mag_students + man_students,
         avg_students = (mag_students + man_students) / 2) %>%
  mutate(`Average students per district observed` = avg_students)

ggplot(data = temp2,
       aes(x = date,
           y = mag_to_man)) +
  geom_point(alpha = 0.2,
             aes(size = `Average students per district observed`)) +
  geom_line(alpha = 0.2) +
  geom_smooth() +
  geom_vline(xintercept = as.numeric(as.Date(c('2015-11-15', '2016-01-15'))), lty = 2, alpha = 0.6, color = 'darkred') +
  theme_maltem() +
  xlab('Date') +
    ylab('Absenteeism: Magude as % of Manhiça')  +
  ggtitle('Magude absenteeism as a % of Manhiça absenteeism') +
  theme(legend.position = 'bottom')
```

In the above chart, daily noise makes comparison difficult. Therefore, we can instead calculate a monthly rolling mean.

```{r}
look_back <- 30
temp2$rolling_mean <- NA
for (i in 1:nrow(temp2)){
  temp2$rolling_mean[i] <-
    (mean(temp2$mag[temp2$date <= temp2$date[i] &
                      temp2$date >= (temp2$date[i] - look_back)], na.rm = TRUE) /
    mean(temp2$man[temp2$date <= temp2$date[i] &
                      temp2$date >= (temp2$date[i] - look_back)], na.rm = TRUE)
       ) * 100
}

# Remove those with NA
temp3 <- temp2[!is.na(temp2$`Average students per district observed`),]

# # At least 30 students from each area
temp3 <- temp3 %>%
  filter(
    man_students >= 10,
         mag_students >= 10,
         # Remove outliers,
         rolling_mean >=20,
         rolling_mean <= 400)
# temp3 <- temp2

ggplot(data = temp3,
       aes(x = date,
           y = rolling_mean)) +
  geom_point(alpha = 0.2,
             aes(size = `Average students per district observed`)) +  
  geom_line(alpha = 0.5) +
  # geom_smooth(aes(weight = `Students under observation`)) +
  #   geom_vline(xintercept = as.numeric(as.Date('2015-11-15')), lty = 2, alpha = 0.6, color = 'darkred') +
  theme_maltem() +
  xlab('Date') +
  ylab('Absenteeism: Magude as % of Manhiça') +
  geom_hline(yintercept = 100) +
    ggtitle('Magude absenteeism as a % of Manhiça absenteeism',
            paste0(look_back, ' day rolling average')) +
  theme(legend.position = 'bottom') +
  geom_vline(xintercept = as.numeric(as.Date(c('2015-11-15', '2016-01-15'))), lty = 2, alpha = 0.6, color = 'darkred')
```

Alternatively, we can explore relative absenteeism (Maugde as a percentage of Manhiça) aggregated at the monthly level.

```{r}
temp4 <- temp2 %>%
  mutate(year_month = as.Date(paste0(format(date, '%Y-%m-'), '01'))) %>%
  group_by(year_month) %>%
  summarise(man = weighted.mean(man, w = man_students, na.rm = TRUE),
            mag = weighted.mean(mag, w = mag_students, na.rm = TRUE)) %>%
    mutate(mag_to_man = (mag / man)  * 100) 

ggplot(data = temp4,
       aes(x = year_month,
           y = mag_to_man)) +
  geom_line(alpha = 0.6, color = 'darkgreen') +
  geom_point(alpha = 0.6, color = 'darkgreen') +
  theme_maltem() +
  xlab('Date') +
  ylab('Absenteeism: Magude as % of Manhiça') +
  geom_vline(xintercept = as.numeric(as.Date(c('2015-11-15', '2016-01-15'))), lty = 2, alpha = 0.6, color = 'darkred') +
  ggtitle('Montly absenteeism: Magude as a % of Manhiça')
```
