
---
title: MALTEM school absenteeism overview
author: 
  - Laia Cirera
  - Joe Brew
  - Elisa Sicuri 
fig_height: 2.6
fig_width: 4
header-includes: \usepackage{longtable}
highlight: zenburn
latex_engine: xelatex
output:
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
tables: yes
toc: yes
---

```{r, echo = FALSE, warning = FALSE, message = FALSE, comment = NA, error= FALSE }
# No scientific notation
options(scipen=999)

#####
# PACKAGES
#####
library(dplyr)
library(xtable)
library(ggplot2)
library(knitr)
library(png)
library(grid)
library(extrafont)
library(tidyr)
library(gridExtra)
library(maptools)
library(rgeos)
library(ggmap)

loadfonts(device = 'postscript', quiet = TRUE)

# BASIC KNITR OPTIONS
opts_chunk$set(comment = NA, echo = FALSE, warning = FALSE, message = FALSE, error = FALSE, fig.height = 3, fig.align='center')
options(xtable.comment = FALSE)
```

```{r, results='hide'}
# GET THEME FOR GGPLOTS
source('theme.R')
# READ IN DATA
source('read_in.R')
```

```{r, fig.align='center', fig.height = 3}
# TITLE MAP
ggplot() +
  coord_map() +
  geom_polygon(data = moz3_fortified,
       aes(x = long, y =lat, group = group), fill = 'grey', color = 'white') +
    geom_polygon(data = maputo_fortified,
       aes(x = long, y =lat, group = group), fill = 'darkred', color = NA, alpha = 0.6) +
  theme_tufte() +
  theme(axis.ticks.length = unit(0.001, "mm")) + labs(x=NULL, y=NULL) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      # legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank())
```


\newpage


# Introduction  

This document presents a largely visual overview of intitial absenteeism data related to the MALTEM program in Magude and Manhiça, Mozambique. Note that we do not present "findings" or "results", but rather general overviews. 

This report is programmed as a "dynamic document"; it can reproduced automatically with updatd data at any point in time.

## Data details  

```{r}
# Get schools locations
schools <- 
  df %>%
  group_by(SCHOOL_NAME) %>%
  summarise(lat = GPS_LAT[1],
            lng = GPS_LNG[1],
            n = length(unique(student_id)))

# Organize number of student days by school
student_days <- 
  df %>%
  group_by(SCHOOL_NAME) %>%
  summarise(n = n())

```


Data have been collected over three months on `r length(unique(df$SCHOOL_NAME))` schools, with a total of `r length(unique(df$student_id))` students. In total, `r sum(student_days$n)` unique student-days have been analyzed.

```{r}

# Map the schools
g1 <- ggplot() +
  coord_map() +
  geom_polygon(data = maputo_fortified,
       aes(x = long, y =lat, group = group), fill = 'white', color = 'grey') +
  theme_tufte() +
  theme(axis.ticks.length = unit(0.001, "mm")) + labs(x=NULL, y=NULL) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      # legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank()) +
  geom_point(data = schools %>%
               rename(Students = n),
             aes(x = lng, y = lat, size = Students),
             col = 'darkred', alpha = 0.3) 

g2 <- 
  ggplot(data = student_days %>% 
           arrange(n) %>%
           mutate(SCHOOL_NAME = gsub('ESCOLA|PRIMARIA|DO|PRIMEIRO E SEGUN|GRAU', '', SCHOOL_NAME)) %>%
           mutate(SCHOOL_NAME = factor(SCHOOL_NAME, levels = SCHOOL_NAME)) %>%
           filter(!is.na(SCHOOL_NAME)),
         aes(x= SCHOOL_NAME, y = n)) +
  geom_bar(stat = 'identity', 
           fill = 'darkred',
           alpha = 0.6) +
    theme_tufte() +
    theme(axis.text.x = element_text(angle = 90, 
                                     hjust = 1, 
                                     size = 6)) +
  xlab('School name') +
  ylab('Student-days')
multiplot(g1, g2, cols = 2)
```


## By time  

When we combine all schools, we can estimate the overall seasonality of absenteeism.

```{r}
by_time <- 
  df %>% group_by(date) %>%
  summarise(n = n(),
            absences = length(which(absence))) %>%
  mutate(p = absences / n * 100)

ggplot(data = by_time,
       aes(x = date, y = p)) +
  geom_point(color = 'darkred', alpha = 0.2) +
  geom_smooth() +
  xlab('Date') +
  ylab('Absenteeism rate') +
  ggtitle('Absenteeism over time')
```


## By district  

Generally speaking, absenteeism is slightly higher in Magude than in Manhiça:

```{r}
# Group by district
cols <- c('blue', 'red')
by_district <-
  df %>%
  group_by(date,district) %>%
  summarise(absent = length(which(absence)),
            n = n()) %>%
  mutate(p = absent/ n * 100)
by_district <- by_district[!is.na(by_district$district),]

ggplot(data = by_district,
       aes(x = date, y = p, group = district,
           color = district)) +
  geom_point(alpha = 0.4) +
  geom_smooth(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  xlab('Date') +
  ylab('Percent absent') +
  ggtitle('School absenteeism: Manhica vs. Magude')  +
  scale_colour_manual(values = cols, name = 'District')

```

However, it is worth noting that the number of students in Magude is fewer, and therefore more subject to natural variance:

```{r}
school_time <- df %>%
  group_by(date, district) %>%
  summarise(n = n())

ggplot(data = school_time,
       aes(x = date, y = n, group = district, color = district)) +
  geom_line() +
  scale_colour_manual(values = cols, name = 'District') +
  xlab('Date') +
  ylab('Students') +
  ggtitle('Number of students under observation by district/time')

```

If we aggregate by month, we can examine absenteeism in a slightly smoother fashion:

```{r}
by_month <- df %>%
  group_by(district, year, month) %>%
  summarise(n = n(),
            absences = length(which(absence))) %>%
  mutate(p = absences / n * 100) %>%
  filter(!is.na(district))

ggplot(data = by_month,
       aes(x = month, y = p, group = district, color = district)) +
  geom_line(size = 2, alpha = 0.6) +
  xlab('Month') +
  ylab('Percent absent') +
  ggtitle('Student absenteeism by month') +
  scale_colour_manual(values = cols, name = 'District')


```

\newpage 

## By age  

There is no clear trend between a student's age and their absenteeism rate.

```{r}
by_age <- df %>%
  mutate(age = as.numeric((Sys.Date() - DOB) / 365.25 )) %>%
  group_by(student_id) %>%
  summarise(n = n(),
            absences = length(which(absence)),
            age = first(age)) %>%
  mutate(p = absences / n * 100) %>%
  filter(age > 5, age < 16)

ggplot(data = by_age,
       aes(x = age, y = p)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  xlab('Age') +
  ylab('Absenteeism rate') +
  ggtitle('Student age and absenteeism')
  
```

## By gender

Generally speaking, females have lower absenteeism than males: 

```{r}
by_gender_date <- df %>%
  group_by(gender, date) %>%
  summarise(n = n(),
            absences = length(which(absence))) %>%
  mutate(p = absences / n * 100) %>%
  ungroup() %>%
  mutate(gender = ifelse(gender == 'Female', 'Girls', 
                         ifelse(gender == 'Male', 'Boys', NA)))
```

```{r}
ggplot(data = by_gender_date,
       aes(x = date, y = p, group = gender, color = gender)) +
  geom_line(alpha = 0.3) + 
  geom_point(alpha = 0.3) +
  geom_smooth() +
  scale_colour_manual(values = cols, name = 'Gender') +
  xlab('Date') +
  ylab('Absenteeism rate') +
  ggtitle('Absenteeism by gender')

```


## By school 


```{r, eval = FALSE}
by_school_date <- 
  df %>%
  group_by(SCHOOL_NAME, date) %>%
  summarise(n = n(),
            absences = length(which(absence))) %>%
    mutate(p = absences / n * 100) %>%
  ungroup %>%
  mutate(SCHOOL_NAME = gsub('ESCOLA|PRIMARIA|DO|PRIMEIRO E SEGUN|GRAU', '', SCHOOL_NAME))

cols <- rainbow(length(unique(by_school_date$SCHOOL_NAME)))
ggplot(data = by_school_date,
       aes(x = date, y = p, group = SCHOOL_NAME, color = SCHOOL_NAME)) +
  geom_line(alpha = 0.5) +
  scale_colour_manual(values = cols, name = 'School') +
  xlab('Date') +
  ylab('Absenteeism rate')

```


There is a surprisingly high level of variation in absenteeism by school. To assess overall absenteeism, we de-aggregate on time, and examine overall absenteeism by school.

```{r, fig.height = 3.5}
by_school <- 
  df %>%
  group_by(SCHOOL_NAME) %>%
  summarise(n = n(),
            absences = length(which(absence)),
            lat = first(GPS_LAT),
            lng = first(GPS_LNG)) %>%
  mutate(p = absences / n * 100) %>% 
  ungroup %>%
  mutate(SCHOOL_NAME = gsub('ESCOLA|PRIMARIA|DO|PRIMEIRO E SEGUN|GRAU', '', SCHOOL_NAME)) %>%
  filter(!is.na(SCHOOL_NAME))

ggplot(data = by_school %>%
         arrange(desc(p)) %>%
         mutate(SCHOOL_NAME = factor(SCHOOL_NAME, levels = SCHOOL_NAME)),
       aes(x = SCHOOL_NAME, y = p)) +
  geom_bar(stat = 'identity', alpha = 0.6, color = 'darkred') +
  theme(axis.text.x = element_text(angle = 90, 
                                     hjust = 1, 
                                     size = 6)) +
  xlab('School') +
  ylab('Absenteeism rate') +
  ggtitle('Absenteeism by school')
```

## By geography

```{r, fig.height = 4.5}
# Satelite-based atlas map (for health care workers)
if('tempmap.RData' %in% dir('/tmp')){
  load('/tmp/tempmap.RData')
} else {
  temp <- get_map(location = coordinates(maputo)[1,],
                maptype = 'satellite',
                zoom = 8)
  save('temp', file = '/tmp/tempmap.RData')
}

# Create a mapper (one row = one absenteeism point)
mapper <- data.frame(lng = NA, lat = NA)
for (i in 1:nrow(by_school)){
  new_rows <- round(by_school$p[i])
  new_mapper <- data.frame(lng = rep(by_school$lng[i], new_rows),
                           lat = rep(by_school$lat[i], new_rows))
  mapper <- rbind(mapper, new_mapper)
}

# Kernel density surfaces
ggmap(temp) +
    # ylim(-25.5, -24.5) +
    coord_cartesian() +
  stat_density2d(data = mapper,
                 aes(x = lng,
                     y = lat,
                     fill = ..level..),
                 geom = 'polygon') +
                 # colour = class,
                 # bins = 30,
            # facet_wrap(~localidade, nrow = 2) +
  scale_fill_gradient(low = 'darkorange',
                      high = 'darkblue',
                      guide = guide_legend(reverse=TRUE), 
                      name = 'Absenteeism index') +
  xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('Absenteeism by space')
```

## By lunch status  

```{r}
by_lunch_school <- df %>%
  group_by(SCHOOL_NAME) %>%
  summarise(lunch = first(LUNCH[!is.na(LUNCH)]),
            lng = first(GPS_LNG),
            lat = first(GPS_LAT)) 
by_lunch <- by_lunch_school %>%
  group_by(lunch) %>% tally %>%
  mutate(lunch = ifelse(lunch == TRUE, 'Offers free lunch', 
                        'Does not offer free lunch'))

ggplot(data = by_lunch, 
       aes(x = lunch, y = n)) +
  geom_bar(stat = 'identity', color = 'darkred', alpha = 0.6) +
  xlab('Lunch status') +
  ylab('Schools') +
  ggtitle('Distribution of schools offering free lunch')
```

Geographically, the fews schools not offering free lunch are clustered together

```{r}
ggplot() +
  coord_map() +
  geom_polygon(data = maputo_fortified,
       aes(x = long, y =lat, group = group), fill = 'white', color = 'grey') +
  theme_tufte() +
  theme(axis.ticks.length = unit(0.001, "mm")) + labs(x=NULL, y=NULL) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      # legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank()) +
  geom_point(data = by_lunch_school %>% filter(!is.na(lunch)) %>%
                 mutate(lunch = ifelse(lunch == TRUE, 'Offers free lunch', 
                        'Does not offer free lunch')),
             aes(x = lng, y = lat, color = lunch),
             alpha = 0.6)

```

## By day of the week

```{r}
temp <- df
temp$dow <- weekdays(temp$date)
by_day <- temp %>%
  group_by(dow) %>%
  summarise(absences = length(which(absence)),
            n = n()) %>%
  mutate(p = absences / n * 100) %>%
  mutate(dow = factor(dow, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))) 
orderer <- data.frame(dow = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'),
                     order = 1:5)
by_day <- 
  by_day %>% left_join(orderer) 
by_day$dow <- factor(by_day$dow, levels = by_day$dow)

ggplot(data = by_day,
       aes(x = dow, y = p)) +
  geom_bar(stat = 'identity') +
  xlab('Day') +
  ylab('Absenteeism rate') +
  ggtitle('Absenteeism by day of week')
```

## Geographical distribution of boys and girls

```{r}
bg <- df %>%
  group_by(SCHOOL_NAME) %>%
  summarise(boys = length(unique(student_id[gender == 'Male'])),
            girls = length(unique(student_id[gender == 'Female'])),
            students = length(unique(student_id)),
            lat = first(GPS_LAT),
            lng = first(GPS_LNG)) %>%
  mutate(`Percent boys` = boys / students * 100)
ggplot() +
  coord_map() +
  geom_polygon(data = maputo_fortified,
       aes(x = long, y =lat, group = group), fill = 'white', color = 'grey') +
  theme_tufte() +
  theme(axis.ticks.length = unit(0.001, "mm")) + labs(x=NULL, y=NULL) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      # legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank()) +
  geom_point(data = bg,
             aes(x = lng, y = lat, color = `Percent boys`, size = students))
```

## By latrine status  


## By school conditions
\newpage  


# Appendix  

## School-specific absenteeism charts   

```{r, fig.height = 8.5}
# Plot schools
unique_schools <- unique(df$SCHOOL_NAME)
unique_schools <- unique_schools[!is.na(unique_schools)]

for (i in 1:length(unique_schools)){
  # Isolate school in question
  school <- unique_schools[i]
  # Subset data
  sub_df <- df[df$SCHOOL_NAME == unique_schools[i],]
  # Group by day and get absenteeism
  sub_df_grouped <- 
    sub_df %>%
    group_by(date) %>%
    summarise(absent = length(which(absence)),
              n = n()) %>%
    mutate(p = absent/ n * 100)
  
  # Get gathered df
  sub_df_gathered <- 
    gather(sub_df_grouped,
           key,
           value, 
           absent:p)
  cols <- c('darkgrey', 'darkred')
  g1 <- 
    ggplot() +
    geom_area(data = sub_df_gathered %>%
                filter(key %in% c('n', 'absent')) %>%
                mutate(key = ifelse(key == 'n', 'Eligible student-days', 
                                    ifelse(key == 'absent', 'Student absences',
                                           key))),
              aes(x = date, y = value, group = key, fill = key),
              alpha = 0.6) +
    xlab('Date') +
    ylab('Students') +
    ggtitle(unique_schools[i])  +
     scale_fill_manual(name="",values=cols)

  
  g2 <- 
    ggplot(data = sub_df_grouped,
           aes(x = date, y = p)) +
    geom_line(alpha = 0.3) +
    geom_point(alpha = 0.6) +
    geom_smooth() +
    ylim(0,100) +
    xlab('Date') +
    ylab('Percent absent')
  
  multiplot(g1, g2)
}
```

```{r, eval = FALSE}


by_school <-
  df %>%
  group_by(date,SCHOOL_NAME) %>%
  summarise(absent = length(which(absence)),
            n = n(),
            district = first(DISTRICT)) %>%
  mutate(p = absent/ n * 100) %>%
  mutate(district = ifelse(district == 1, 'Manhica',
                           ifelse(district == 2, 'Magude',
                                  NA)))

ggplot(data = by_school,
       aes(x = date, y = p, group = district,
           color = district)) +
  geom_point(alpha = 0.3) +
  xlab('Date') +
  ylab('Percent absent') +
  ggtitle('School absenteeism: Manhica vs. Magude
          (each dot is a school-day)
          (lines are district average)') +
  theme_bw() +
  geom_line(data = by_district,
            aes(x = date, y = p, group = district,
                color = district),
            size = 1.4)


# Time map
dates <- sort(unique(df$date))
# start in may
dates <- dates[dates > '2015-05-01']

for (i in 1:length(dates)){
  date <- dates[i]
  formatted_date <- format(date, '%B %d, %Y')
  sub_data <- df[df$date == date,]
  sub_data_grouped <- sub_data %>%
    group_by(SCHOOL_NAME) %>%
    summarise(lat = first(GPS_LAT),
              lng = first(GPS_LNG),
              n = n(),
              a = length(which(absence))) %>%
    mutate(p = a / n * 100)
  plot(maputo)
  points(sub_data_grouped$lng,
         sub_data_grouped$lat,
         cex = sub_data_grouped$p / 10,
         pch = 16,
         col = adjustcolor('darkred', alpha.f = 0.6))
  legend('left',
         pch = 16,
         col = adjustcolor('darkred', alpha.f = 0.6),
         pt.cex = c(5, 10, 25, 50)/ 10,
         legend = c(5, 10, 25, 50),
         title = 'Percent absent')
  title(main = formatted_date)
  Sys.sleep(0.1)
}
```
