---
title: "MALTEM school map and distances from CISM"
author: "Joe Brew, Laia Cirera, Elisa Sicuri"
fig_width: 4
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
fig_height: 2.6
---
```{r, echo = FALSE, warning = FALSE, message = FALSE, comment = NA, error= FALSE, cache = FALSE}
# No scientific notation
options(scipen=999)

# Packages 
library(xtable)
library(knitr)
library(ggplot2) 
# library(dplyr)
library(rgeos)
library(RColorBrewer)
library(Rmisc)
# Libraries
library(rgdal)
library(raster)
library(sp)
library(leaflet)
library(ggmap)
library(readxl)
library(dplyr)
library(readr)
library(geosphere)

# Basic knitr options
opts_chunk$set(comment = NA, 
               echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, # Render report, even with errors
               cache = F)
```

```{r setup, include=FALSE, echo = FALSE}
library(knitr)
knit_engines$set(asis = function(options) {
  if (options$echo && options$eval) knit_child(text = options$code)
})
```


```{r, results='hide'}
# GET THEME FOR GGPLOTS
source('theme.R')
# READ IN DATA
source('read_and_clean_all.R')

# Combine manhica and magude
combined_map <-
  rbind(magude, manhica)

# Fortify
combined_map_fortified <-
  fortify(combined_map, region = 'NAME_3')

# Spatial versoin of schools
schools <- ungroup(schools)
schools_spatial <- schools
schools_spatial$x <- schools_spatial$lng
schools_spatial$y <- schools_spatial$lat
schools_spatial <- schools_spatial[which(!is.na(schools_spatial$lat)),]
coordinates(schools_spatial) <- ~x+y
proj4string(schools_spatial) <- proj4string(moz)
```

# MAPS

## Interactive map

```{r}
cols <- colorRampPalette(brewer.pal(n = 9, name = 'Spectral'))(2)
colors <- ifelse(schools_spatial@data$district_number == '01', 
                 cols[1],
                 cols[2])
leaflet(schools_spatial) %>%
  # addProviderTiles('Esri.WorldImagery') %>%
  addProviderTiles('Stamen.Toner') %>%
  # addProviderTiles('Stamen.TonerLabels') %>%
  addProviderTiles('Esri.NatGeoWorldMap') %>%
  # fitBounds(lng1 = min(coordinates(combined_map)[,1]),
  #           lat1 = min(coordinates(combined_map)[,2]),
  #           lng2 = max(coordinates(combined_map)[,1]),
  #           lat2 = max(coordinates(combined_map)[,2])) %>%
  addCircleMarkers(#lng = schools$lng,
             #lat = schools$lat,
             popup = schools$SCHOOL_NAME,
             col = colors,
             opacity = 0,
             fillOpacity = 0.7)
  
```

## Static map

```{r, fig.align='center', fig.height = 3}
ggplot() +
  coord_map() +
    geom_polygon(data = combined_map_fortified,
       aes(x = long, y =lat, group = group), 
       fill = NA, 
       color = 'black', 
       alpha = 0.6) +
  geom_point(data = schools_spatial@data,
             aes(x = lng,
                 y = lat),
             color = colors) +
  theme(axis.ticks.length = unit(0.001, "mm")) + labs(x=NULL, y=NULL) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      # legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank()) 
```

# Distances

The below table shows the distance from the CISM to each school, using Haverstine's "great-cirlce-distance" (aka, "as the crow flies") in kilometers.

As of November 9, 2016, the average liter of gasoline in Mozambique costs approximately 0.60 euros. With a "wear-and-tear" cost of approximately 0.20 euros per kilometer, a fuel efficiency of approximately 7 kilometers per liter, and a Haverstine inflation factor of 2 (ie, the true road distance is approximately twice the straight-line distance), we can calculate round trip transportation cost for each school.

```{r}
# Define cism location
cism <- data.frame(name = 'CISM',
                   x = 32.803869,
                   y = -25.408462)
cism$lng <- cism$x; cism$lat <- cism$y
coordinates(cism) <- ~x+y
proj4string(cism) <- proj4string(schools_spatial)
distances <- geosphere::distm(x = cism,
                              y = schools_spatial, 
                              fun = distVincentySphere)
distance_results <- data.frame(SCHOOL_NAME = schools_spatial@data$SCHOOL_NAME,
                               distance = round(as.numeric(distances) / 1000, digits = 2))
# Join to schools
schools <- schools %>%
  left_join(distance_results)

schools$euros <- 
  round(((schools$distance * 2) / 7) * (0.6 + 0.2), digits = 2)

x <- schools %>% dplyr::select(SCHOOL_NAME, distance, euros, n_students, district_number)
names(x) <- toupper(names(x))
DT::datatable(x)
```

Note that in the above table, two schools are not listed. We do not have coordinates for these 2 schools:

```{r}
cat(paste0(schools$SCHOOL_NAME[is.na(schools$lat)], collapse = ' | '))
```

However, if we project the _average_ cost of the other schools onto these two, we come up with a total transportation cost for going to all 25 schools of `r round((25 * mean(schools$euros, na.rm = TRUE)), digits = 2)` euros.